\section{Motivation}
\label{sec:benchmarks_motivation}

As discussed in \cref{ch:introduction}, developing agents that can operate in open-world environments requires the ability to detect, respond to, and recover from novelty. Evaluating whether an agent possesses these capabilities, however, is far from straightforward. The unbounded nature of novelty means that agents cannot be tested against a fixed, closed set of scenarios. Instead, evaluation environments must support the systematic introduction of diverse novelties while accommodating agents with fundamentally different architectures and representations.

Existing environments for training and evaluating AI agents have largely been designed under closed-world assumptions. Standard reinforcement learning benchmarks such as OpenAI Gym~\citep{brockman2016openai} and Atari~\citep{bellemare2013arcade} provide well-defined tasks with fixed dynamics, offering no mechanism for injecting environmental changes after training. Planning benchmarks based on PDDL~\citep{mcdermott1998pddl} similarly assume that the domain specification is complete and correct. While some environments support procedural generation or domain randomization, these techniques focus on generalization through diversity rather than on evaluating an agent's response to specific, structured changes in its environment.

Several challenges arise when attempting to evaluate novelty-aware agents. First, the notion of novelty is inherently agent-relative: what constitutes a novelty for one agent may not be novel for another, depending on each agent's perceptual capabilities, knowledge representation, and prior experience. Second, comparing agents across paradigms is difficult when environments are designed with a particular approach in mind. A planning agent requires symbolic state descriptions and operator definitions, whereas a reinforcement learning agent requires numeric observations and reward signals. An evaluation environment must support both kinds of interfaces without favoring one over the other. Third, the metrics commonly used to evaluate agents, such as cumulative reward or task success rate, are insufficient for characterizing how an agent responds to novelty. Metrics must capture not only whether the agent eventually succeeds, but also how quickly it detects the novelty, how severely its performance degrades, and how efficiently it adapts.

This chapter presents two benchmark environments that address these challenges. The first, NovelGridworlds~\citep{goel2021novelgridworlds}, introduces a benchmark with a rich taxonomy of novelties across multiple tasks, together with evaluation metrics for both planning and reinforcement learning agents. The second, NovelGym~\citep{goel2024novelgym}, extends and generalizes this work into a flexible ecosystem with a formal framework for environment transformations, a modular architecture for developing hybrid agent systems, and adaptation-focused evaluation metrics. Together, these environments provide the infrastructure upon which the adaptive methods presented in subsequent chapters are developed and evaluated, and they contribute to the broader research community by offering standardized benchmarks for open-world novelty. Source codes and documentation for both environments are publicly available, and they have been designed to be extensible so that future researchers can build upon them to explore new types of novelty and adaptation strategies.