\section{Related Work}
\label{sec:benchmarks_related_work}

This section reviews prior work on evaluation environments and benchmarks relevant to the study of agents under open-world novelty. The discussion is organized around three themes: standard reinforcement learning and planning benchmarks that operate under closed-world assumptions, environments that introduce variability or non-stationarity, and frameworks specifically designed for open-world novelty evaluation.

\paragraph{Closed-world evaluation benchmarks.}
The dominant paradigm for evaluating sequential decision-making agents assumes fixed, fully specified environments. Reinforcement learning benchmarks such as OpenAI Gym~\citep{brockman2016openai} and the Arcade Learning Environment~\citep{bellemare2013arcade} provide standardized tasks with known dynamics, enabling comparison of learning algorithms under controlled conditions. Similarly, classical planning benchmarks expressed in PDDL~\citep{mcdermott1998pddl} assume that the domain model is complete and correct, and evaluate planners on plan quality and computational efficiency. These environments have been productive for advancing algorithms within their respective paradigms, but they do not support evaluation of how agents respond to changes that violate their learned or specified models---precisely the setting formalized in \cref{sec:open_world_environments}.

\paragraph{Generalization and domain transfer.}
A separate line of work addresses generalization across environments through domain randomization~\citep{tobin2017domain,peng2018sim} and procedurally generated environments. These approaches expose agents to a distribution of tasks during training, with the goal of improving robustness to variation at deployment. While effective for improving generalization within the training distribution, they do not explicitly evaluate how agents respond to post-training changes that introduce mismatch between the agent's models and the environment. The variability is incorporated into the training distribution rather than introduced as a structured, diagnosable deviation, and consequently these approaches do not exercise the adaptation capabilities that are central to open-world operation.

\paragraph{Out-of-distribution detection and novelty detection.}
Research on out-of-distribution (OOD) detection addresses the problem of identifying when an input differs from the training distribution. Methods such as confidence-based detection~\citep{hendrycks2017baseline}, open-set recognition~\citep{scheirer2013toward,bendale2016towards}, and autoencoder-based anomaly detection~\citep{abati2019autoregression} have been studied extensively in perception and classification settings, where novelty is defined in terms of input statistics. However, these methods focus on detection rather than adaptation: they identify that something has changed but do not provide mechanisms for the agent to modify its behavior in response. In the framework of \cref{sec:novelty}, detection addresses only one aspect of the agent's response to mismatch; the benchmarks introduced in this chapter must additionally support evaluation of how agents recover from and adapt to detected novelties.

\paragraph{Open-world novelty frameworks.}
The DARPA SAIL-ON program~\citep{senator2019science} catalyzed research on agents operating in environments with unanticipated novelty, motivating formal definitions of open-world novelty~\citep{boult2021towards,langley2020open} and agent architectures for novelty handling~\citep{muhammad2021novelty,goel2024neurosymbolic}. Within this broader effort, NovGrid~\citep{balloch2022novgrid} provides a flexible gridworld platform for studying novelty, supporting configurable environment parameters and novelty injection. The Polycraft World AI Lab~\citep{goss2023polycraft} offers a richer, Minecraft-based environment for evaluating novelty detection and accommodation at scale. NovelCraft~\citep{feeney2022novelcraft} provides a dataset and benchmark for visual novelty detection and discovery in open worlds.

These environments represent important steps toward open-world evaluation, but they address different aspects of the problem than the benchmarks introduced in this chapter. NovGrid focuses on environment configurability but does not provide a formal taxonomy that connects novelty types to their impact on agent task performance, as developed in \cref{sec:novelty_taxonomy}. The Polycraft environment supports large-scale evaluation but is tightly coupled to a specific domain and agent interface. NovelCraft targets visual novelty detection rather than the full cycle of detection, adaptation, and recovery.

\paragraph{Gap addressed by this chapter.}
The benchmarks introduced in this chapter---NovelGridworlds and NovelGym---address gaps that cut across these lines of work. They provide environments that support controlled injection of diverse classes of mismatch, organized according to the taxonomy of novelty by domain element and task impact developed in \cref{sec:novelty}. They accommodate agents across multiple paradigms---symbolic planning, reinforcement learning, and hybrid architectures---through flexible observation interfaces. And they introduce evaluation metrics that capture not only task success but the dynamics of adaptation: how quickly performance degrades, how efficiently the agent recovers, and whether beneficial novelties are exploited. These capabilities are essential for evaluating the claims developed in \cref{sec:abstractions}: that structured abstractions over actions and interactions determine whether agents can adapt efficiently and generalize effectively under novelty.