\section{Summary and Future Directions}
\label{sec:benchmarks_summary}

This chapter presented two benchmark environments for developing and evaluating agents that operate under open-world novelty. NovelGridworlds introduced a structured novelty taxonomy,
% cross-paradigm evaluation through a socket interface,
and metrics that capture both performance degradation and novelty detection. NovelGym extended this foundation with a formal environment representation, composable novelty transformations, a modular architecture supporting hybrid planning-learning agents, and an evaluation protocol with adaptation-focused metrics. Together, these environments provide the experimental infrastructure for the adaptive methods developed in the subsequent chapters of this dissertation.

The experimental results across both environments reveal consistent themes. Planning agents detect novelties reliably through discrepancies between expected and observed state transitions but lack mechanisms for adapting their models in response. Reinforcement learning agents can potentially adapt through continued interaction but require substantially more experience and are brittle when novelties alter the observation space or introduce catastrophic states. Hybrid agents that integrate planning with learning-based recovery consistently achieve faster adaptation, as measured by $T_\text{adapt}$, because they decompose the problem and focus learning on the specific subgoal affected by the novelty. However, the failure-driven nature of hybrid adaptation has a blind spot: beneficial novelties that do not cause execution failures, such as the chest shortcut, go undetected and unexploited.

The running example of the axe novelty---in which trees become unbreakable unless the agent acquires and uses a new tool---illustrates these tensions concretely. A planning agent detects the failure of the {\tt break} operator but cannot learn the recovery behavior. A reinforcement learning agent can eventually learn to use the axe but requires orders of magnitude more experience. A hybrid agent localizes the failure and learns a targeted recovery policy. In \cref{ch:rapid_learn}, we develop the RAPid-Learn framework, which formalizes and extends this hybrid recovery mechanism, using the axe novelty and other NovelGym scenarios as primary evaluation domains.

Several directions for future work emerge from the benchmarking framework introduced in this chapter.

\paragraph{Multi-agent open-world settings and LLM-based agentic systems.} NovelGym's architecture, built on PettingZoo, already supports multi-agent environments, but the evaluation in this chapter focused on single-agent scenarios with passive dynamic entities. Extending the framework to fully multi-agent open-world settings would open a qualitatively different research frontier. In multi-agent environments, novelty is no longer limited to environmental transformations: it can originate from other agents---a teammate that changes strategy, a collaborator that acquires new capabilities, or an adversary that adopts previously unseen behaviors. The transformation formalism introduced in \cref{sec:ng_novelty} naturally extends to such settings by defining agent-behavioral transformations alongside environmental ones, and the evaluation metrics can be generalized to capture coordination robustness and collective adaptation.

This extension becomes particularly timely given the rapid emergence of LLM-based multi-agent systems~\citep{wang2024survey}. Architectures such as AutoGPT, BabyAGI, and multi-agent collaboration frameworks~\citep{ahn2022can,huang2022language} are increasingly being deployed in open-ended settings where agents must plan, delegate, and recover from failures autonomously. These systems bring broad world knowledge and flexible reasoning capabilities, but they also face distinctive challenges under novelty: hallucination of infeasible actions, brittleness to unexpected tool or API changes, and difficulty maintaining coherent plans when the environment or collaborators behave in unanticipated ways. Despite the rapid proliferation of such systems, there is no standardized benchmark for evaluating their open-world novelty handling capabilities. NovelGym's composable novelty framework, cross-paradigm agent support, and adaptation-focused metrics provide the groundwork for building such a benchmark. Extending the ecosystem to support LLM-based agent interfaces---and designing novelty scenarios that specifically target the reasoning, grounding, and coordination challenges faced by agentic LLM systems---would establish a rigorous evaluation infrastructure for this emerging and rapidly growing class of embodied agents.

\paragraph{Goal-directed curiosity for novelty handling.} The experimental results in this chapter reveal that curiosity-driven exploration, as implemented through ICM~\citep{pathak2017curiosity}, provides an imperfect signal for novelty adaptation. The forward model's prediction error correlates with state novelty but not necessarily with task relevance: many novel states are irrelevant to the agent's current goal, and the intrinsic reward disperses exploration rather than focusing it on the specific transitions needed for recovery. This limitation is not unique to ICM; other intrinsic motivation methods, including Random Network Distillation~\citep{burda2019exploration} and count-based exploration~\citep{bellemare2016unifying}, similarly reward visitation of unfamiliar states without regard to goal relevance.

Recent work by \citet{lorang2025curiosity} takes a promising step toward addressing this limitation by combining ICM-driven exploration with a symbolic operator learner and reward machines derived from planning in an ``imaginary'' symbolic domain. In their approach, undirected curiosity drives the agent toward novel states, and when those states produce new symbolic transitions, the system abstracts them into planning operators and uses them to construct Linear Temporal Logic reward machines that densify the reward signal. This dual-channel architecture---broad exploration via ICM, structured exploitation via symbolic planning---demonstrates faster adaptation than prior hybrid methods in a robotic manipulation domain with sequential novelty injections. However, the curiosity signal itself remains undirected: the ICM still rewards all prediction errors uniformly, and the authors' own ablations confirm that ICM alone continues to suffer from exploring irrelevant states when the novelty is difficult to model or when sensory feedback is uninformative.

This points to a still-open research direction: developing goal-conditioned or task-aware curiosity mechanisms for open-world settings~\citep{liu2022goal}. Rather than leaving curiosity undirected and compensating with a separate exploitation channel, the exploration bonus itself could be shaped by the agent's current subgoal or by the discrepancy between planned and observed operator effects---directing curiosity toward the {\em functionally relevant} aspects of the novelty from the outset. In the context of hybrid planning-learning systems, the symbolic planner provides a natural source of goal structure: when an operator fails, the planner identifies the specific preconditions or effects that were violated, and curiosity could be directed toward states that resolve precisely those discrepancies. Such an approach would bridge the gap between undirected exploration (which is sample-inefficient even when combined with reward machines) and purely failure-driven adaptation (which misses beneficial novelties). The benchmark environments introduced in this chapter, with their structured novelties spanning beneficial, detrimental, and irrelevant categories, provide a natural testbed for evaluating such goal-centric exploration strategies.

\paragraph{Toward physically grounded open-world simulation.} The environments presented in this chapter operate in discrete gridworlds, which provide the controlled abstraction necessary for systematic evaluation but lack the physical complexity of real-world robotic settings. Recent work on open-world simulation platforms, such as Virtual Community~\citep{zhou2025virtual}, demonstrates the feasibility of physics-grounded, multi-agent environments that combine robotic manipulation, human-robot interaction, and large-scale scene generation in realistic 3D worlds. Bridging the gap between the structured novelty injection framework of NovelGym and the physical fidelity of such simulation platforms is a promising direction. This would enable evaluation of how novelty-aware agents cope with continuous state spaces, contact dynamics, and perceptual uncertainty---challenges that are central to the robotic interaction methods developed in Chapter~5 of this dissertation.