\subsection{Hybrid Planning and Learning}
\label{sec:hybrid}

Symbolic planning and reinforcement learning provide complementary approaches to sequential decision-making. Symbolic planning enables structured reasoning over high-level representations but assumes accurate and complete models of the environment. Reinforcement learning enables agents to learn behaviors through interaction but typically operates over lower-level or subsymbolic state representations and may lack explicit structure. In many settings, reinforcement learning can also be viewed hierarchically, where high-level symbolic operators correspond to abstract actions, and their execution is realized through lower-level policies or planners. Hybrid planning and learning systems aim to combine these paradigms by integrating symbolic reasoning with learned or executed behaviors.

In hybrid systems, decision-making is defined over two representations. The symbolic model is represented by a planning task $\PlanningTask = \langle \Fluents, \Operators, \SymState_0, \Goal \rangle$,and the execution is formalized by a Markov decision process $M = \langle \States, \Actions, \Transition, \Reward, \Discount \rangle$. The symbolic model operates over discrete representations, whereas the execution model operates over environment states that may be continuous or stochastic. An abstraction function $\Abstraction : \States \rightarrow \SymStates$ maps environment states to symbolic states.

Operators in the symbolic model correspond to executable behaviors in the environment. Each operator $\Operator \in \Operators$ is associated with a policy $\Policy_{\Operator}(a \mid s)$ that defines its execution in the environment. A symbolic plan $\Plan = [\Operator_1, \dots, \Operator_n]$ therefore induces a sequence of policies executed in the environment.

A hybrid planning and learning system is defined as $\Hybrid = \langle \PlanningTask, M, \Abstraction \rangle$, which formalizes the separation between symbolic decision-making over abstract representations and execution in the environment.
