\subsection{Reinforcement Learning}
\label{sec:rl}

Reinforcement learning models sequential decision-making in which an agent learns to select actions through interaction with an environment so as to maximize cumulative reward. At each time step, the agent observes a state, selects an action, receives a reward signal, and transitions to a new state.

Reinforcement learning problems are modeled as a Markov Decision Process (MDP), defined as $M = \langle \States, \Actions, \Transition, \Reward, \Discount \rangle$, where $\States$ is the set of states, $\Actions$ is the set of actions, $\Transition(s' \mid s,a)$ defines the probability of transitioning to state $s'$ after taking action $a$ in state $s$, $\Reward(s,a)$ is the reward function, and $\Discount \in [0,1)$ is the discount factor.

At time step $t$, the agent is in state $s_t \in \States$, selects an action $a_t \in \Actions$, receives reward $r_{t+1} = \Reward(s_t,a_t)$, and transitions to a new state $s_{t+1}$ according to $\Transition(\cdot \mid s_t,a_t)$. The Markov property assumes that the transition and reward depend only on the current state and action.

A policy $\Policy(a \mid s)$ defines the behavior of the agent by specifying a probability distribution over actions given a state. The objective of reinforcement learning is to find a policy that maximizes the expected cumulative reward.

The return from time step $t$ is defined as $G_t = \sum_{k=0}^{\infty} \Discount^k r_{t+k+1}$. The state-value function is defined as $\Value^\Policy(s) = \Expect_\Policy [G_t \mid s_t = s]$, and the action-value function is defined as $\QValue^\Policy(s,a) = \Expect_\Policy [G_t \mid s_t = s, a_t = a]$.

These value functions satisfy recursive relationships known as the Bellman expectation equations. The state-value function satisfies
\[
\Value^\Policy(s) = \Expect_\Policy \left[ r_{t+1} + \Discount \Value^\Policy(s_{t+1}) \mid s_t = s \right],
\]
and the action-value function satisfies
\[
\QValue^\Policy(s,a) = \Expect_\Policy \left[ r_{t+1} + \Discount \Expect_{a' \sim \Policy(\cdot \mid s_{t+1})} \QValue^\Policy(s_{t+1}, a') \mid s_t = s, a_t = a \right].
\]

In discrete settings, the state and action spaces are finite, i.e., $|\States| < \infty$ and $|\Actions| < \infty$. In this case, value functions such as $\Value^\Policy(s)$ and $\QValue^\Policy(s,a)$ can be represented explicitly in tabular form, and reinforcement learning algorithms can directly estimate these quantities for each state or state-action pair.

In continuous settings, the state space $\States \subseteq \Real^n$ and/or the action space $\Actions \subseteq \Real^m$ may be continuous. In such cases, value functions and policies cannot be represented exactly and are instead approximated using parameterized function approximators, typically implemented using neural networks. 
A policy is represented as $\Policy_{\theta}(a \mid s)$, where $\theta$ denotes the parameters of the policy, and value functions are represented as $\Value_{\phi}(s)$ and $\QValue_{\phi}(s,a)$, where $\phi$ denotes the parameters of the function approximator.

Learning in continuous settings involves optimizing the parameters $\theta$ and $\phi$ to maximize expected return. These formulations are particularly important in robotics and control tasks, where states and actions are naturally continuous and policies must operate over high-dimensional spaces.
