\section{Motivation}
\label{sec:gen_motivation}

The preceding chapters addressed how agents can detect, evaluate, and recover from novelty in open-world environments. \Cref{ch:benchmarks} introduced benchmark environments and metrics for studying agent behavior under mismatch, and \cref{ch:rapid_learn} developed hybrid planning--learning methods that exploit action abstractions to achieve efficient adaptation when execution fails. Both contributions operate within the framework established in \cref{ch:foundations}: novelty arises from mismatch between the agent's internal models and the environment, and structured abstractions determine whether this mismatch can be managed effectively.

This chapter addresses the second dimension of the dissertation's central claim: that structured abstractions over interactions enable learned behaviors to \emph{generalize} across variations in objects, physical parameters, and robotic embodiment. While \cref{ch:rapid_learn} showed that action abstractions support efficient adaptation \emph{after} failure, the work in this chapter asks a complementary question: can interaction abstractions be designed so that learned skills remain valid \emph{before} failure occurs---that is, so that a policy trained on one object or robot transfers to new objects and robots without retraining?

The challenge is rooted in how robotic manipulation policies represent their interaction with the world. As discussed in \cref{sec:interaction_abstractions}, the standard approach in robot learning defines policies over robot-centric representations: joint angles, end-effector poses, or motor torques. These representations tightly couple the learned behavior to a specific embodiment and control interface. A policy trained to open a drawer with a Panda robot encodes not just the task-relevant forces and motions, but also the Panda's specific kinematics, joint limits, and control dynamics. The result is a policy that solves the task on the training platform but fails when the robot changes, the object changes, or the physical parameters (mass, friction, dimensions) change---even when the underlying task physics remain the same.

This limitation is not merely a practical inconvenience. It reflects a representational failure: the policy's state and action spaces do not separate the aspects of the interaction that are invariant across variations (the forces and motions required to manipulate the object) from those that are contingent on the training instance (the robot's kinematics, the specific object's dimensions, the particular friction coefficient). In the language of \cref{sec:abstractions}, the representation fails to abstract away the agent, and therefore does not capture the natural complexity of the task~\citep{konidaris2019necessity}.

An important clarification is in order regarding the relationship between generalization and novelty. The variations studied in this chapter---different objects, different robots, different physical parameters---are anticipatable in the sense that the representation is explicitly designed to be invariant to them. This differs from the unanticipated novelties addressed in \cref{ch:rapid_learn}, where the agent encounters changes it could not have foreseen. However, the connection to the dissertation's central claim is direct: the right choice of abstraction can \emph{preempt} certain classes of mismatch entirely, ensuring that variations which would constitute novelty under a robot-centric representation are absorbed by the invariant structure of the force-space representation without any adaptation being required. In this sense, interaction abstractions and action abstractions play complementary roles: interaction abstractions eliminate the need for adaptation across anticipated variation, while action abstractions provide the machinery for recovery when genuinely unanticipated changes occur. Together, they define a spectrum of responses to open-world variation, from preemptive invariance to reactive adaptation.

This chapter develops an alternative: \emph{object-centric force-space representations} for robotic manipulation. The core idea is to define policies that operate in the space of forces applied to the object, conditioned on object-relative state descriptions, rather than in the space of robot commands. By representing interaction at the level of the object's dynamics rather than the robot's control, the learned behavior captures the physical invariants of the task---what forces to apply, where on the object to apply them, and what motion results---while remaining agnostic to the specific robot that executes them. This constitutes a structured interaction abstraction in the sense of \cref{sec:interaction_abstractions}: the representation is organized around the causally relevant aspects of the manipulation task, creating a buffer between the learned policy and the variability of objects and embodiments.

We instantiate this idea through two complementary contributions that address different classes of sustained-contact manipulation:

\begin{itemize}[leftmargin=*]
\item \textbf{FLEX: Force-based learning for articulated object manipulation} (\cref{sec:flex}). FLEX learns object-centric force policies for manipulating articulated objects---drawers, doors, microwaves, dishwashers---whose motion is constrained by prismatic or revolute joints~\citep{fang2025flex}. Policies are trained entirely in simulation without modeling any robot, using only the object's joint configuration and the applied force as state and action. An interactive joint parameter estimation procedure enables deployment on physical robots, and an operational space controller translates the learned forces into robot commands. FLEX demonstrates generalization across unseen objects of the same joint type and direct transfer to three different robotic platforms (Kinova, Panda, UR5) without retraining.

\item \textbf{Pushing heavy objects: Force policies for rigid body transport} (\cref{sec:pho}). This work extends the object-centric force-space approach to the transport of rigid objects along prescribed paths under unknown mass and friction conditions~\citep{goel2025pho}. A policy is trained in simulation to output planar boundary forces conditioned on a normalized contact-resistance descriptor, with domain randomization over mass and friction. A lightweight impedance-based probing routine estimates the interaction regime at deployment, enabling zero-shot sim-to-real transfer. Experiments on a Boston Dynamics Spot mobile manipulator demonstrate robust transport of heavy payloads (up to 12.5~kg) along curved trajectories.
\end{itemize}

Together, these contributions demonstrate that object-centric force-space representations provide a practical and principled pathway to generalizable manipulation. FLEX shows that the approach generalizes across objects and robots for articulated manipulation; pushing heavy objects shows that it generalizes across physical parameters and transfers zero-shot from simulation to a legged mobile manipulator for rigid body transport. Both instantiate the interaction abstraction framework developed in \cref{sec:interaction_abstractions} and provide empirical evidence for the claim that structured abstractions over interactions enable generalization under the kind of variation that characterizes open-world environments.