\section{Summary and Future Directions}
\label{sec:gen_summary}

This chapter presented two contributions that demonstrate how object-centric force-space representations enable learned manipulation behaviors to generalize across objects, physical parameters, and robotic embodiments. FLEX learns force-based policies for articulated object manipulation that transfer directly to three different robot platforms without retraining, achieving over an order-of-magnitude improvement in training efficiency compared to robot-centric baselines. The pushing heavy objects framework extends this approach to rigid body transport under unknown mass and friction, demonstrating zero-shot sim-to-real transfer on a legged mobile manipulator.

Both contributions instantiate the interaction abstraction framework developed in \cref{sec:interaction_abstractions}. The key design principle is the same in both cases: by defining policies in the space of forces applied to the object---conditioned on object-relative state descriptions---the learned behavior captures the physical invariants of the task while remaining agnostic to the specific robot, object instance, or physical parameters. This constitutes a structured interaction abstraction that serves as a buffer between the learned policy and the variability of the world, consistent with the unifying perspective of \cref{sec:unifying_perspective}.

The parallel with \cref{ch:rapid_learn} is instructive. In that chapter, action abstractions enabled efficient adaptation by localizing failure to specific operators and focusing learning on the affected component. In this chapter, interaction abstractions enable generalization by organizing the policy's representation around task-relevant physical invariants rather than embodiment-specific details. Both forms of abstraction reduce the scope of what must be learned or relearned when conditions change---but they operate at different levels: action abstractions structure the \emph{sequence} of behaviors, while interaction abstractions structure the \emph{representation} of each behavior.

Several limitations and future directions emerge from this work.

\paragraph{Richer interaction descriptors.} The pushing heavy objects framework conditions on a single scalar $z$ capturing the effective contact resistance. This is sufficient for quasi-static transport under uniform friction but cannot represent spatially varying friction, anisotropic contact, or time-varying resistance. Extending the descriptor to a richer representation---potentially learned from interaction history---would broaden the range of applicable scenarios.

\paragraph{Dynamic manipulation.} Both FLEX and the pushing framework assume quasi-static interaction regimes. Extending to dynamic manipulation, where inertial effects dominate and contact modes switch rapidly, would require richer state representations and potentially hybrid quasi-static/dynamic control formulations.

\paragraph{Contact recovery and robustness.} The dominant failure mode in real-world experiments was contact loss, either from slip at the gripper--object interface or from changes in contact geometry during body repositioning. Incorporating contact-state estimation from wrist force/torque sensing and developing reactive re-grasping strategies would improve robustness in unstructured environments.

\paragraph{Integration with planning and adaptation.} The force-space policies developed in this chapter operate at the execution level. Integrating them as executors within the hybrid planning--learning framework of \cref{ch:rapid_learn} would create a system that can both plan over abstract task structure and execute through generalizable force-based skills. When a novel object is encountered whose physical properties differ from prior experience, the planning layer could detect the mismatch and invoke targeted adaptation---using the interaction descriptor estimation as a lightweight mechanism for characterizing the new regime---while the force-space representation ensures that the adapted behavior transfers across embodiments. This integration represents a natural convergence of the action and interaction abstractions studied in this dissertation.

\paragraph{Multi-contact and collaborative manipulation.} Both frameworks consider single-contact manipulation. Extending to multi-contact settings---where multiple forces are applied simultaneously, or where multiple robots collaborate---would require coordinating force policies across contact points and managing the interactions between them.