\chapter{Background and Preliminaries}

This chapter introduces the formal frameworks that underpin the methods developed throughout this dissertation. The goal is not to provide a comprehensive survey of the fields, but rather to establish the mathematical models and notation required to describe the proposed algorithms and systems.

The dissertation builds on three core paradigms for sequential decision making: symbolic planning, reinforcement learning, and hybrid planning and learning. These paradigms provide complementary capabilities for reasoning, learning, and adaptation in complex environments.

Section~\ref{sec:symbolic_planning} introduces the formalism of symbolic planning, including representations of states, predicates, and action operators. Section~\ref{sec:rl} presents reinforcement learning using the Markov Decision Process framework. Section~\ref{sec:hybrid} introduces hybrid planning and learning architectures that integrate symbolic reasoning with learning-based decision making, and formalizes the notion of open-world environments and novelty.

These formalisms provide the foundation for the methods presented in subsequent chapters. Following this formal background, later sections of this chapter situate the thesis within the broader literature on open-world learning, adaptive agents, and robotics.
\input{chapters/chapter2/chapter2_background/symbolic_planning}
\input{chapters/chapter2/chapter2_background/reinforcement_learning}
\input{chapters/chapter2/chapter2_background/hybrid_planning_learning}
