\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Despite significant advances in artificial intelligence and robotics, and recent progress toward more flexible and open-world capable systems, many approaches still exhibit brittle behavior when confronted with changes beyond their training distributions, reflecting a continued reliance on closed-world assumptions. In real-world environments, new objects, altered dynamics, and unforeseen task variations arise naturally, and even small changes can render existing knowledge inadequate, exposing a limitation in how current systems represent and reason about their interaction with the world.

This dissertation advances the study of open-world agent–environment interaction by developing the formal, algorithmic, and representational foundations required for agents to operate under novelty. We formalize novelty as a mismatch between an agent’s internal models and the environment and introduce environment models, transformations, and evaluation methodologies for its systematic study. 
Building on this foundation, the dissertation shows how abstractions shape learning and behavior in open-world settings. Symbolic abstractions over actions and tasks expose the structure of execution, enabling failures to be localized and converted into targeted learning problems for rapid adaptation. 
Object-centric force-space abstractions represent interaction in terms of object dynamics rather than robot-specific control, enabling generalization in robotic manipulation across objects, embodiments, and physical conditions.

Viewed as a whole, this dissertation introduces a formal framework for open-world agent–environment interaction and develops the tools to study and address novelty within this setting. It highlights how abstractions enable adaptation and generalization under changing conditions, providing a step toward artificial intelligence and robotic systems that can operate more robustly in real-world environments.