\section{Motivation}
\label{sec:rl_motivation}

This dissertation investigates learning, adaptation, and generalization for autonomous agents and robots operating in open-world environments. In such settings, agents must act under the assumption that their models of the world are incomplete: changes in the environment can introduce novelties that invalidate previously learned dynamics, action models, or state representations. 

\Cref{ch:benchmarks} addressed a key prerequisite for studying this problem by introducing standardized testbeds and evaluation metrics for agent behavior under novelty. The NovelGridworlds and NovelGym benchmarks provide modular environments that support diverse classes of novelty, accommodate agents across multiple paradigms, and evaluate not only task success but also the dynamics of adaptation. These include how quickly an agent detects deviations from expected outcomes, the extent to which performance degrades following a novelty, and how efficiently the agent recovers to achieve its goals. With this evaluation infrastructure in place, we now turn to the central methodological question: \emph{how should an agent adapt its behavior when the underlying model of the environment is no longer valid?}

The difficulty of this problem arises from fundamental limitations of existing paradigms. In a symbolic planning framework, actions are defined by operators with preconditions and effects, and plans are generated under the assumption that these models are correct. When a novelty alters the environment dynamics, execution may fail because the expected effects of an operator do not materialize. While such failures can be detected, the planner itself has no mechanism to construct new operators or modify executors to account for the change. 

Reinforcement learning methods, in contrast, operate directly over state-action interactions and can, in principle, learn new behaviors in response to such changes. However, in the absence of structural guidance, learning must proceed through broad exploration over the state-action space, often resulting in high sample complexity and the loss of previously acquired knowledge. In open-world settings characterized by abrupt, task-relevant changes, this leads to inefficient adaptation and poor reuse of prior experience.

The central insight motivating this chapter is that these two paradigms offer complementary strengths. Symbolic planning provides a structured representation of the task that can identify where execution has failed—specifically, which operator’s preconditions or effects are violated—thereby localizing the source of the novelty. Reinforcement learning, on the other hand, provides a mechanism for acquiring new behaviors through interaction, enabling the agent to learn executors that restore the validity of the plan. By combining these capabilities, adaptation can be framed as a targeted process: learning is invoked only at points of failure, and prior knowledge is preserved wherever the model remains valid.

To illustrate, consider the axe novelty introduced in \cref{sec:benchmarks_motivation}. An agent that has learned to craft items by breaking trees encounters a change in which the environment dynamics are altered such that trees can no longer be broken directly, but instead require the use of an axe. The agent generates a plan that includes the \texttt{break} operator, whose preconditions are satisfied, but whose expected effects fail to occur during execution. This execution failure signals a mismatch between the agent’s model and the environment. To recover, the agent must incorporate newly relevant entities and actions, and learn a new executor that enables progress toward the goal. Importantly, this adaptation should occur without relearning the entire task. A purely symbolic system cannot synthesize the required behavior, and a purely learning-based system must rediscover large portions of the task. In contrast, a hybrid approach can use the structure of the failed plan to localize the adaptation problem and learn only the missing capability required to restore successful execution.

This chapter develops methods and architectures for adaptive planning and learning under such conditions. First, \cref{sec:ipt,sec:rapid_learn_method} introduce \emph{RAPid-Learn} (Learning to Recover and Plan Again)~\citep{goel2022rapidlearn}, a framework that integrates symbolic planning with reinforcement learning to enable efficient recovery from execution failures. RAPid-Learn formalizes the notion of an \emph{integrated planning task}, characterizes execution failures as instances of missing or invalid executors, and employs knowledge-guided exploration to learn recovery policies with substantially fewer interactions than standard reinforcement learning. 

Second, \cref{sec:architecture} presents a neurosymbolic cognitive architecture~\citep{goel2024neurosymbolic} that embeds RAPid-Learn within a broader system for novelty detection, characterization, and accommodation. This architecture integrates symbolic reasoning, subsymbolic perception, and reinforcement learning to support the full lifecycle of adaptation, from detecting violations of expected operator effects to acquiring and reusing new behaviors under persistent novelty. \Cref{sec:ch4_evaluation} evaluates both contributions: RAPid-Learn as a standalone method on the NovelGridworlds benchmark, and the integrated architecture on a large-scale evaluation spanning 288 novelty scenarios.