\section{Related Work}
\label{sec:rl_related_work}

This section reviews prior work related to the adaptation methods developed in this chapter. The discussion is organized around four themes: hierarchical and temporal abstraction in reinforcement learning, hybrid planning and learning systems, transfer learning and continual adaptation, and prior approaches to novelty handling in open-world settings.

\paragraph{Hierarchical reinforcement learning and temporal abstraction.}
The executor formalism used in this chapter draws on the options framework~\citep{sutton1999options}, which extends reinforcement learning to include temporally extended actions defined by initiation sets, policies, and termination conditions. Options enable agents to reason and learn at multiple levels of temporal granularity, and have been shown to improve exploration and transfer in hierarchical settings. The connection between temporal abstraction and symbolic planning has been formalized by \citet{konidaris2018skills}, who show that the symbols necessary and sufficient for planning can be derived from the initiation sets and effect distributions of a set of skills, and by \citet{konidaris2016constructing}, who develops the skill-symbol loop in which new skills give rise to new symbolic representations. The executor discovery problem introduced in this chapter can be viewed as the inverse of the skill-symbol loop: given a symbolic operator whose executor has failed, the agent must learn a new skill (executor) that satisfies the operator's specification.

\paragraph{Hybrid planning and learning.}
Integrating symbolic planning with learned execution has been explored in several contexts. Task and motion planning (TAMP) systems~\citep{garrett2021integrated,kaelbling2011hierarchical} combine symbolic task planning with continuous motion planning, but typically assume complete and correct domain models. Plan monitoring~\citep{fritz2007monitoring} and plan repair~\citep{fox2006plan} address execution failures within the symbolic planning paradigm, either by detecting when a plan becomes suboptimal or by modifying a plan to account for unexpected states. However, these approaches operate entirely at the symbolic level and cannot acquire new low-level behaviors when the failure stems from a mismatch between the domain model and the environment dynamics. The SPOTTER framework~\citep{sarathy2021spotter} takes a step toward integrating planning with learning by using reinforcement learning to extend symbolic planning operators, but it focuses on operator acquisition in a known environment rather than recovery from novelty-induced execution failures. RAPid-Learn differs from these approaches in that it explicitly targets the executor discovery problem: when an operator fails due to novelty, it instantiates a focused reinforcement learning problem guided by the symbolic structure of the failed plan.

\paragraph{Transfer learning and continual adaptation.}
When an agent encounters a changed environment, one natural approach is to transfer knowledge from the original task to the new setting. Policy reuse~\citep{glatt2017policy} transfers a learned policy and selects between the old and new policies during execution. Actor-critic transfer methods~\citep{clegg2017haptics} transfer both the policy and value function networks. Progressive neural networks~\citep{rusu2016progressive} avoid catastrophic forgetting by freezing previously learned columns and adding new capacity for new tasks, while elastic weight consolidation~\citep{kirkpatrick2017overcoming} protects important weights from modification. These methods address the problem of leveraging prior experience, but they treat adaptation as a monolithic learning problem over the full state-action space. When novelty affects only a specific component of the task, this global approach is inefficient: the agent must relearn or readjust its entire policy rather than targeting the specific behavior that failed. The experimental results in \cref{sec:rl_eval_rapid} confirm this limitation, showing that transfer learning baselines require one to two orders of magnitude more experience than RAPid-Learn to achieve comparable post-novelty performance.

\paragraph{Open-world novelty handling.}
The DARPA SAIL-ON program~\citep{senator2019science} motivated the development of agent architectures for detecting and accommodating novelty in open-world environments. \citet{boult2021towards} and \citet{langley2020open} provide formal frameworks for reasoning about novelty, while \citet{muhammad2021novelty} present an agent architecture that combines novelty detection with reactive accommodation strategies. \citet{balloch2023neuro} develop neuro-symbolic world models that adapt internal representations in response to detected novelty. The cognitive architecture presented in \cref{sec:architecture} extends this line of work by integrating symbolic and neural novelty detection with a cascading recovery pipeline that progresses from lightweight symbolic strategies to learning-based executor discovery, and by evaluating on a substantially larger and more diverse set of novelty scenarios than prior work.

\paragraph{Gap addressed by this chapter.}
Across these lines of work, a common limitation is the absence of a principled mechanism for converting a global task failure into a targeted learning problem. Symbolic methods can detect and localize failures but cannot synthesize new behaviors. Learning methods can acquire new behaviors but lack the structural guidance needed for efficient adaptation. The RAPid-Learn framework addresses this gap by leveraging the action abstractions developed in \cref{sec:action_abstractions}---the operator--executor decomposition---to bridge symbolic failure diagnosis with focused reinforcement learning, enabling adaptation that is both targeted and sample-efficient.