\section{Thesis Outline}
\label{sec:outline}

The remainder of this dissertation is organized as follows.

Chapter 2 presents the foundations of open-world agent--environment interaction. It formalizes the notion of novelty as a mismatch between an agent's internal models and the environment, and reviews prior work in perception, planning, learning, and adaptation under such conditions. The chapter identifies key challenges that arise when agents operate under incomplete and evolving knowledge, and motivates the need for representations that support effective behavior under novelty.

Chapter 3 introduces benchmark environments, formalism, and evaluation methodologies for systematically studying novelty. These environments enable controlled injection of diverse classes of mismatch and provide metrics for analyzing how agent performance degrades and recovers under changing conditions. This chapter establishes an experimental foundation for evaluating adaptation and generalization in open-world settings.

Chapter 4 investigates adaptation under novelty. It presents hybrid planning--learning methods that enable agents to recover from execution failures by leveraging structured action abstractions. These approaches localize failures to specific components of a plan and enable targeted learning, allowing agents to efficiently acquire missing behaviors without relearning entire tasks.

Chapter 5 addresses generalization in robotic manipulation. It presents object-centric, force-based learning approaches that represent interaction at the level of object dynamics. These representations enable learned skills to generalize across objects with similar semantics, across variations in physical properties, and across different robotic platforms, supporting robust performance under diverse conditions.

Chapter 6 concludes the dissertation by summarizing the main findings and discussing future directions for research on autonomous agents and robots operating in open-world environments