Autonomous agents and robots, whether operating in homes, warehouses, or unstructured outdoor environments, share a common operational structure: they perceive their surroundings, reason and plan over possible actions, and execute those actions in the physical world. This perceive-plan-execute loop, and its natural extension to include recovery and adaptation, underlies decades of work in artificial intelligence and robotics. Under closed-world assumptions, where the environment, objects, and dynamics are known in advance, each stage of this loop has been studied extensively and with considerable success. Classical perception pipelines reliably detect known object categories. Symbolic planners generate provably correct plans over well-specified domain models. Learned controllers execute manipulation and locomotion skills in environments that match their training distribution.

However, when agents are deployed in open-world environments, the assumptions underlying each stage of this loop are violated. New objects may appear that were not present during training. Environmental dynamics may change in ways that invalidate learned models. Actions that previously succeeded may fail due to unforeseen circumstances. The central challenge for open-world autonomy is not any single stage in isolation, but the fact that novelty can arise at any point in the loop and propagate failures across stages: an unrecognized object leads to an incorrect plan, which leads to an execution failure, which requires recovery that the agent may not be equipped to perform.

This section reviews prior work relevant to each stage of the perceive--plan--execute loop under open-world conditions. Rather than providing exhaustive coverage of any single area, the discussion is organized around three pillars of open-world autonomy---sensing and perception, planning and learning, and execution and recovery---and aims to identify the key challenges and limitations that motivate the contributions of this dissertation.

\input{chapters/chapter2/background/related_work/open_world_sensing_rw.tex}
\input{chapters/chapter2/background/related_work/open_world_planning_learning_rw.tex}
\input{chapters/chapter2/background/related_work/open_world_execution_recovery_rw.tex}
\input{chapters/chapter2/background/related_work/summary_rw.tex}