Some efforts have sought to address multiple stages of the open-world agent loop in an integrated manner. World models \citep{ha2018world, hafner2023mastering} offer a potential unifying framework by learning environment dynamics that can serve both perception and planning, though they typically do not handle discrete, qualitative changes in the environment. End-to-end foundation models for robotics \citep{pmlr-v229-zitkovich23a, team2024octo} implicitly combine perception, reasoning, and action selection in a single model, but their behavior under genuine novelty is not well understood. Task and motion planning systems \citep{garrett2021integrated, kaelbling2011hierarchical} integrate symbolic and geometric reasoning but inherit the brittle model assumptions of classical planning. Despite growing interest in open-world autonomy, the field remains largely fragmented: methods that address perception under novelty, planning under incomplete models, and recovery from execution failures are typically developed in isolation from one another.

The preceding discussion has examined open-world autonomy 
through the lens of the perceive--plan--execute loop, 
identifying challenges at each stage: sensing under novelty, 
planning and learning with incomplete models, and execution 
and recovery when prior knowledge becomes insufficient. 
Drawing primarily from the challenges identified in planning 
and learning (Section~\ref{sec:rw_planning_learning}) and 
execution and recovery (Section~\ref{sec:rw_execution}), this 
dissertation makes progress on three interconnected challenges. 
First, the lack of standardized environments for systematically 
evaluating agents under controlled novelty limits rigorous 
comparison of approaches and slows progress in open-world AI. 
Second, existing hybrid planning-learning systems do not 
adequately address novelty handling: they lack mechanisms for 
detecting execution failures caused by qualitative environmental 
changes, learning recovery behaviors, and reusing acquired 
knowledge across subsequent encounters with novelty. Third, 
robotic skill learning methods remain limited in their ability 
to produce interaction policies that generalize across objects 
and platforms, particularly for contact-rich manipulation tasks.

These three challenges correspond to the core contributions of 
this dissertation. Chapter~3 introduces benchmark environments 
for evaluating open-world agents under controlled novelty 
injection. Chapter~4 develops hybrid planning-learning 
architectures that enable agents to adapt and recover when 
previously learned models become insufficient. Chapter~5 
presents object-centric, force-based reinforcement learning 
methods for robotic manipulation that generalize across objects 
and robot embodiments. Together, these contributions advance 
the goal of building autonomous agents and robots that can 
operate robustly in open-world environments where novelty is 
not the exception but the norm.