\section{Summary of Contributions}
\label{sec:conclusion_contributions}

\paragraph{Evaluation frameworks for open-world novelty.}
\Cref{ch:benchmarks} introduced NovelGridworlds and NovelGym, benchmark environments that enable the systematic study of agent behavior under novelty. These environments support controlled injection of diverse classes of mismatch---including changes to objects, environment dynamics, and task structure---organized according to the novelty taxonomy developed in \cref{sec:novelty_taxonomy}. They accommodate agents across multiple paradigms (symbolic planning, reinforcement learning, and hybrid architectures) through flexible observation interfaces, and provide evaluation metrics that capture the dynamics of adaptation: how quickly performance degrades following a change, and how efficiently the agent recovers. These benchmarks established the experimental foundation on which the subsequent contributions were evaluated.

\paragraph{Action abstractions for efficient adaptation under novelty.}
\Cref{ch:rapid_learn} developed RAPid-Learn, a hybrid planning--learning framework that enables agents to recover from execution failures caused by novelty. The key insight is that structured action abstractions---the operator--executor decomposition formalized in \cref{sec:action_abstractions}---allow failures to be localized to specific components of a plan. When an executor fails because the environment dynamics have changed, the framework converts a global task failure into a targeted learning problem: the agent acquires only the new executor needed to restore successful execution, without relearning previously valid behaviors. Experimental results demonstrated one to two orders of magnitude improvement in sample efficiency compared to unstructured reinforcement learning and transfer learning baselines. The cognitive architecture further extended this framework with a cascading recovery pipeline, progressing from lightweight symbolic strategies to learning-based adaptation, evaluated across a substantially larger and more diverse set of novelty scenarios.

\paragraph{Interaction abstractions for generalizable skill learning.}
\Cref{ch:generalization} introduced object-centric, force-based representations for robotic manipulation that capture interaction at the level of object dynamics rather than robot-specific control. FLEX demonstrated that policies learned entirely in force space transfer directly to three different robotic platforms without retraining, achieving over an order-of-magnitude improvement in training efficiency compared to robot-centric baselines. The pushing heavy objects framework extended this approach to rigid body transport under unknown mass and friction, demonstrating zero-shot sim-to-real transfer on a legged mobile manipulator. Both contributions instantiate the interaction abstraction framework of \cref{sec:interaction_abstractions}: by defining policies in the space of forces applied to the object, the learned behavior captures the physical invariants of the task while remaining agnostic to the specific robot, object instance, or physical parameters.