\section{Future Directions}
\label{sec:future_directions}

The contributions of this dissertation open several avenues for future research. We organize these directions around three themes: extending the technical contributions, pursuing their integration, and connecting to the broader challenges of open-world autonomy.

\paragraph{Scaling structured abstractions to richer domains.}
The benchmark environments and adaptation methods developed in this dissertation were evaluated in gridworld and game-based domains that, while supporting diverse classes of novelty, operate at a level of complexity below real-world deployment settings. Extending the hybrid planning--learning framework to domains with continuous state spaces, partial observability, and richer task structure would test the scalability of the action abstraction approach. The force-space methods in \cref{ch:generalization} operate in continuous robotic manipulation settings, but on a restricted class of tasks (articulated objects and planar pushing). Extending to multi-step manipulation sequences---where force-based skills must be composed into longer-horizon plans---would exercise both the action and interaction abstractions simultaneously.

\paragraph{Integrating perception with adaptation and generalization.}
As discussed in \cref{sec:rw_sensing}, this dissertation assumes access to task-relevant state information and does not address the perception problem directly. In real-world deployment, the agent must also perceive the environment, detect perceptual novelties, and determine their relevance to the current task. Integrating open-world perception systems---such as open-vocabulary detectors and vision-language models---with the adaptation and generalization mechanisms developed here would close an important loop in the agent architecture. A particularly interesting question is whether the novelty taxonomy of \cref{sec:novelty_taxonomy} can be extended to perceptual novelties, and whether the structured abstraction framework can be applied to the perception pipeline itself.

\paragraph{Bridging structured abstractions and foundation models.}
\Cref{sec:foundation_models} argued that the abstraction-based approach developed in this dissertation and the foundation model paradigm are complementary. Realizing this complementarity in practice requires developing concrete architectures that combine the two. One promising direction is using large language models to initialize or refine symbolic domain models, with the hybrid framework providing the machinery for detecting when these models are incorrect and adapting accordingly. Another is using VLA models as initial executors within the operator--executor framework, with the targeted learning mechanisms of \cref{ch:rapid_learn} enabling efficient fine-tuning when the initial executor fails. A third is augmenting foundation model-based agents with force-space representations for contact-rich manipulation, grounding their behavior in physical interaction dynamics.

\paragraph{Formal connections between novelty type and recovery strategy.}
The novelty taxonomy of \cref{sec:novelty_taxonomy} classifies novelties by their impact on task performance (prohibitive, obstructive, beneficial, nuisance), and the adaptation framework of \cref{ch:rapid_learn} provides a cascading recovery pipeline. However, the connection between novelty type and the appropriate recovery strategy is currently heuristic: the agent tries lightweight strategies first and escalates as needed. Developing formal principles for selecting recovery strategies based on diagnosed novelty type---for example, proving that nuisance novelties never require executor relearning, or that prohibitive novelties always require model revision---would provide stronger theoretical foundations for adaptation under novelty.

\paragraph{Multi-agent and collaborative open-world settings.}
The framework developed in this dissertation considers a single agent operating in an environment. In many real-world settings, multiple agents must coordinate in the presence of novelty---for example, a team of robots in a warehouse where new objects or procedures are introduced. Extending the novelty taxonomy, benchmark environments, and adaptation mechanisms to multi-agent settings raises new challenges: novelty may affect agents asymmetrically, coordination strategies may need to be revised, and the interaction between agents' adaptation processes must be managed. The force-space representations in \cref{ch:generalization} provide a natural starting point for multi-agent manipulation, where multiple robots must coordinate their forces on a shared object.

\paragraph{Lifelong learning and knowledge accumulation.}
The adaptation methods in \cref{ch:rapid_learn} enable an agent to recover from individual novelty events, but they do not address the broader problem of lifelong learning: accumulating knowledge over many encounters with novelty and leveraging this accumulated experience to handle future changes more efficiently. An agent that has recovered from several different types of novelty should be better prepared for the next one---for example, by recognizing patterns in how novelties manifest or by maintaining a library of previously learned executors that can be reused or adapted. Developing mechanisms for such cumulative learning within the structured abstraction framework is an important direction for building agents that improve over extended deployment.